---
title: "ITMD514_05  Lab 1"
author: "Surajit Patra"
date: ""
output: html_document
---


```{r}
# We will require the tidyverse library. If you do not already have it installed type the following command to do so in the Console; or follow RStudio prompts to install
# install.packages("tidyverse")
# The following loads the library for our use
library(tidyverse) 
```

### 1. Changing the author field and file name.

##### (a) Change the `author:` field on the Rmd document from Your Name Here to your own name.

##### (b) Rename this file to "lab01_ClassAndSectionNumber_LastName_FirstName_.Rmd", where your info is updated accordingly; e.g. "lab01_STAT514_01_Smith_John.Rmd
### 2.  Hello World!

The following R code chunk prints the text 'Hello world!'.

```{r}
print("Hello world!")
```

#### (a) Modify the code chunk below to print your name
```{r}
print("Surajit Patra")
```

<br>

### 3.  Creating sequences
The `c()` operator allows you to form a vector from its arguments.  If we're trying to build a vector containing a sequence of numbers, there are several useful functions at our disposal.  These are the colon operator `:` and the sequence function `seq()`.

##### `:` Colon operator: 
```{r}
1:19 # Numbers 1 to 10
394:415 # Numbers 394 to 415
```

##### `seq` function: `seq(from, to, by)`
```{r}
seq(1,25,1) # All integers from 1 to 25
seq(1,25,2) # Odd integers from 1 to 25
seq(2,25,2) # Even integers from 2 to 25
```

> To learn more about a function, type `?functionname` into your console.  E.g., `?seq` pulls up a Help file with the R documentation for the `seq` function.

#### (a) Use the operator `:` to output the sequence of numbers from 3 to 48
```{r}
3:48
```

#### (b) Use the `seq()` method to output the sequence of numbers from 3 to 48 in increments of 3
```{r}
seq(3,48,3)
```

#### (c) Save the sequence from (a) as a variable `x`, and the sequence from (b) as a variable `y`.  Output their product `x*y`
```{r}
x <- 3:48
y <- seq(3, 48, 3)
print(x*y)
```

<br>

### 4. Cars data

First basic example of data frames and plotting.

In this example you will explore a very simple, classic dataset.  The `cars` data comes with the default installation of R. Let us find out more about this dataset.
```{r}
?(cars)
```
So we can see this is a data frame, with 50 observations on 2 variables. To peek into the few observations use the command `head(cars)`.

```{r}
head(cars)
```

To access the variables speed and dist of cars, we can use `cars[,1]', or `cars$speed'.

```{r}
speed_cars <- cars[,1]
dist_cars <- cars[,2]
```

One could also use the `attach()` command to allow access to `speed` and `dist` as though they were vectors in one's own workspace. Use `?attach' to read about why this is not good practice.

```{r}
?attach
```

#### (a) Calculate the average, variance, and standard deviation of the speed and distance variables of the cars dataset, and save them in appropriately named variables.
```{r}
# average distance of cars
average_dist <- mean(dist_cars)
print(average_dist)

# average speed of cars
average_speed <- mean(speed_cars)
print(average_speed)

# variance distance of cars
variance_dist <- var(dist_cars)
print(variance_dist)

# variance speed of cars
variance_speed <- var(speed_cars)
print(variance_speed)

# standard deviation of distance of cars
standarddev_dist <- sd(dist_cars)
print(standarddev_dist)

# standard deviation of speed of cars
standarddev_speed <- sd(speed_cars)
print(standarddev_speed)
```

<br>

<br>  We will produce a histogram of stopping distance using the `qplot` function. (Note that this method is deprecated, but we will use it in this first assignment due to its simplicity)
```{r}
qplot(dist_cars) # Histogram of stopping distance
```

The `qplot(x,y,...)` function can also be used to plot a vector `y` against a vector `x`.  As always, type `?qplot` into the Console to learn more about the basic qplot function. 

#### (b) Use the `qplot(x,y)` function to create a scatterplot of dist_cars against speed_cars.
```{r}
qplot(dist_cars, speed_cars)
```